---
title: "Figure 3"
date: "2025-4-29"
editor_options: 
  chunk_output_type: console
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Loading
```{r}
### Please download the required files for Figure 3 from dropbox: https://www.dropbox.com/scl/fo/ovijw9p59q9cwsh3y2ykj/ADem9wbI-rBJtZn6UhF0f_E?rlkey=ehil1bxzefx3knezhbjsocdc9&e=1&st=d6erhd7y&dl=0
```

## Libraries
```{r}
suppressPackageStartupMessages({
  # Load required packages
  library(rlang)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(scran)
  library(ggthemr)
  library(ClassifyR)
  library(survival)
  library(survminer)
  library(stringr)
})

minMaxScaler <- function(x) {
  (x-min(x, na.rm=T))/(max(x, na.rm=T)-min(x, na.rm=T))
}
rename <- dplyr::rename
select <- dplyr::select
slice <- dplyr::slice

# Set ggplot theme
ggthemr_reset()
ggthemr('pale')
mycolors <- c("#999999", "#0072B2", "#E69F00", "#F0E442", "#D9B3FF", "#009E73",  
               "#D55E00", "#5D8AA8", "#CC79A7", "#56B4E9",
              "#F3B3A6", "#A5AB81", "#B2182B", "#4393C3", "#CDBE6B", 
              "#80CDC1", "#F4A582", "#BABABA", "#CCEBC5", "#DECBE4",
              "#FDDFDF", "#B3DE69", "#FDBF6F", "#CCECE6", "#FB8072")

dl_method_pal <- c(
  "THItoGene" ="#BA7FB5", 
  "DeepPT"="#80B1D2", 
  "DeepSpaCE"="#F18072",
  "ST-Net"="#8CD0C3",
  "HisToGene"="#f1c232",
  "Hist2ST"="#F9B063",
  "GeneCodeR"="#BCB9D8",
  "iStar"= "#B3D46B",
  "GHIST" = "#B2182B"
  )

mypalette <- define_palette(
  swatch = mycolors,
  gradient = c(lower = mycolors[1L], upper = mycolors[3L])
)

ggthemr(mypalette)
theme_update(panel.grid.major = element_line(linetype="dotted"))

th <-   theme(text=element_text(size=16),
                axis.text.x = element_text(angle = 0, hjust = 0.5),
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                panel.background = element_rect(colour = "black", size=0.7, fill=NA))

```


## Data

Load data for HER2ST dataset

```{r}
# Load data
# Get metrics
pred_feat_cor <- readRDS("./her2st_pred_feat_cor_8.rds")

# Get GHIST metrics
ghist_pred_feat_cor <- readRDS("./ghist_pred_feat_cor.rds") %>%
  mutate(model_id = case_when(model_id == "Proposed" ~ "GHIST",
    TRUE ~ model_id
  ))
# Combine calculated metrics of comparison methods with GHIST
pred_feat_cor <- rbind(pred_feat_cor, ghist_pred_feat_cor)
```

*pred_feat_cor* is a dataframe with the following columns:

- img_id: id of image
- model_id: the name of method predicting gene expression
- gene: name of the gene that was predicted
- pred_type: whether the evaluation metrics were calculated on the "train", "test" or 
"validation" sets
- train_fold: an id for the fold the model was trained and tested on
- cor_pearson: Pearson correlation
- cor_spearman: Spearman correlation
- var_exprs: variance of the gene expression data the model was trained on
- var_exprs_orig: variance of the ground truth gene expression counts
- var_pred: variance of the predicted gene expression
- mean_exprs: mean of the gene expression data the model was trained on
- mean_exprs_orig: mean of the ground truth gene expression counts
- mean_pred: mean of the predicted gene expression
- rmse: root mean squared error
- mi: mutual information
- js_div: Jensen-Shannon divergence
- nrmse_range: RMSE normalised using the range of the data
- nrmse_sd: RMSE normalised using the standard deviation of data
- ssim: structural similarity index
- auc_n: original ground truth gene expression is binarised by $GE <= n$ and $GE > n$. 
auc_n represents the auc of the predicted gene expression distinguishing the binarisation of the original gene expression

```{r}
glimpse(pred_feat_cor)
```

## Load SVG
```{r}
svg_genes <- readRDS("./her2st_SVG_filter_top20.rds") 
```

### All metrics, all data

#### Figure 3a & b

Boxplot/violin plots of the average PCC and SSIM between the ground truth gene expression and predicted gene expression. Metrics measured from the test fold of a 4-fold CV, averaged over each gene across HER2+ ST datasets.

```{r fig.height=5.5, fig.width=10}
#' Plot Boxplots of Correlations
pred_gene_msr_df <- pred_feat_cor %>%
  mutate(dataset = "her2+") %>%
  filter(pred_type == "test") %>%
  select(gene, pred_type, train_fold, model_id, dataset,
         img_id, cor_pearson, ssim) %>%
  # Average iamges over each patient
  group_by(gene, pred_type, train_fold, model_id, dataset) %>%
  summarise_if(is.numeric, function(col) mean(col,na.rm = T)) %>%
  # # Average all folds and patients over each gene
  group_by(pred_type, gene, model_id, dataset) %>%
  summarise_if(is.numeric, function(col) mean(col,na.rm = T))

th <-   theme(text=element_text(size=15),
                axis.text.x = element_text(angle = 0, hjust = 0.5),
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                panel.background = element_rect(colour = "black", size=0.7, fill=NA) )

pred_gene_msr_df <- pred_gene_msr_df%>%
  arrange(factor(model_id, levels=c("GHIST", "iStar", "THItoGene", "Hist2ST", "DeepPT", "DeepSpaCE", "GeneCodeR", "HisToGene", "ST-Net")))
pred_gene_msr_df$model_id <- factor(pred_gene_msr_df$model_id, levels = unique(pred_gene_msr_df$model_id))

fig_3a_b <- pred_gene_msr_df %>%
  pivot_longer(cols=c("cor_pearson", "ssim"),
               names_to="metric",
               values_to="value") %>%
  mutate(metric = factor(metric, levels = c("cor_pearson", "ssim"))) %>%
  mutate(metric=case_when(
    metric=="cor_pearson" ~ "PCC",
    metric=="ssim" ~ "SSIM",
    TRUE ~ "other"
  ) %>%
    factor(., levels=c("PCC", "SSIM"))) %>%
  ggplot() +
  aes(x=model_id, y=value, fill=model_id, col=model_id) +
  geom_violin(alpha=0.2) +
  geom_boxplot(alpha=0.5, width=0.3) +
  facet_wrap(~metric, nrow=1, scales="free") +
  coord_flip()+
  scale_color_manual(values=dl_method_pal)+
  scale_fill_manual(values=dl_method_pal)+
  th + 
  labs(x="", y="")

fig_3a_b
```

### Results for top genes

Top five correlated genes in the HER2+ ST dataset

```{r}
# Get top genes in her2+
top_cor_df <- pred_feat_cor %>%
  filter(pred_type=="test") %>% 
  mutate(dataset="her2+") %>%
  group_by(gene, model_id, dataset) %>%
  summarise(mean_cor = mean(cor_pearson, na.rm=T),.groups="drop") %>%
  pivot_wider(names_from = "model_id", values_from="mean_cor") %>%
  rowwise() %>% 
  mutate(overall_mean_cor = mean(c_across(-c(gene, dataset)))) %>%
  arrange(dataset,-overall_mean_cor) %>%
  ungroup()
```

#### Figure 3c
```{r fig.height=3.5, fig.width=5}
# Plot top genes her2+
plotTopGenesBar <- function(top_cor_df) {
  top_cor_df %>%
    group_by(dataset) %>%
    dplyr::slice(1:5) %>%
    mutate(gene = factor(gene, levels=rev(gene))) %>%
    pivot_longer(cols=!c("gene", "overall_mean_cor", "dataset"),
                 names_to="model_id",
                 values_to="cor")  %>% 
    ggplot() +
    aes(x=gene,y=cor,col=model_id,fill=model_id)+
    geom_bar(stat="identity", position=position_dodge(),
             alpha=0.6, width=0.75) +
    geom_text(data=. %>%
                distinct(gene, overall_mean_cor),
              aes(x=gene, 
                  label = round(overall_mean_cor,2), 
                  y=Inf),
              size = 5, 
              hjust=1.1,
              inherit.aes = F)+
    facet_wrap(~dataset, scales="free")+
    scale_y_continuous(expand = c(0, Inf)) +
    scale_color_manual(values=dl_method_pal)+
    scale_fill_manual(values=dl_method_pal)+
    coord_flip()+
    labs(y="Mean PCC", x="")
}

th <-   theme(text=element_text(size=16),
                axis.text.x = element_text(angle = 0, hjust = 0),
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                panel.background = element_rect(colour = "black", size=0.7, fill=NA) )

p_topgenes_her2 <- plotTopGenesBar(top_cor_df %>%
                                     mutate(dataset = ifelse(dataset=="her2+", "HER2+ ST", dataset))) +   theme(plot.title = element_text(size = 5, face = "bold"),
        legend.position="bottom",
        legend.text = element_text(size=8),
    legend.key.size = unit(12, 'points'),
    legend.title = element_text(size=9),
    strip.text.x = element_text(size=16)) +
  labs(col="",fill="") +
  th

p_topgenes_her2
```

#### Figure 3d & e
```{r}
#get gene subgroups
pred_gene_msr_df <- pred_feat_cor %>%
  mutate(gene_set = "All Genes") %>%
  bind_rows(
    pred_feat_cor %>%
      ungroup() %>%
      semi_join(svg_genes, by = c("gene", "img_id")) %>%
      mutate(gene_set = "SVGs")
  ) %>%
  select(gene, pred_type, train_fold, model_id,
         img_id, cor_pearson, ssim, gene_set) %>%
  # Average iamges over each patient
  group_by(gene, pred_type, train_fold, gene_set, model_id) %>%
  summarise_if(is.numeric, function(col) mean(col,na.rm = T)) %>%
  # Average all folds and patients over each gene
  group_by(pred_type, gene, gene_set, model_id) %>%
  summarise_if(is.numeric, function(col) mean(col,na.rm = T))

pred_gene_msr_df <- pred_gene_msr_df%>%
  arrange(factor(model_id, levels=c("GHIST", "iStar", "THItoGene", "Hist2ST", "DeepPT", "DeepSpaCE", "GeneCodeR", "HisToGene", "ST-Net")))
pred_gene_msr_df$model_id <- factor(pred_gene_msr_df$model_id, levels = unique(pred_gene_msr_df$model_id))

fig_SVG <- pred_gene_msr_df %>%
  filter(pred_type == "test") %>%
  filter(gene_set %in% c("SVGs")) %>%
  pivot_longer(cols=c("cor_pearson","ssim"),
               names_to="metric",
               values_to="value") %>%
  mutate(metric = factor(metric, levels = c("cor_pearson", "ssim"))) %>%
  mutate(metric=case_when(
    metric=="cor_pearson" ~ "PCC",
    metric=="ssim" ~ "SSIM",
    TRUE ~ "other"
  ) %>%
    factor(., levels=c("PCC","SSIM"))) %>%
  ggplot() +
  aes(x=model_id, y=value, fill=model_id, col=model_id) +
  geom_violin(alpha=0.2) +
  geom_boxplot(alpha=0.5, width=0.3) +
  facet_wrap(~metric, nrow=1, scales="free") +
  coord_flip()+
  scale_color_manual(values=dl_method_pal)+
  scale_fill_manual(values=dl_method_pal)+
  theme(legend.position = "none") +
  th + 
  labs(x="", y="")

fig_SVG
```

### Survival Analysis on HER2+ TCGA 
#### Data
```{r}
# load the dataset
tcga_rnaseq_data <- readRDS(
  file="./tcga_breast_matched_rnaseq_sce.rds"
)
```


```{r}
surv_risk_classifyr_list_top5feat_cv_3fold100rep <- readRDS(
  file="./top20.rds")

surv_eval_dat <- surv_risk_classifyr_list_top5feat_cv_3fold100rep

surv_risk_classifyr_cindex <- lapply(
  seq_along(surv_eval_dat),
  function(i) {
    surv_eval_dat[[i]]@predictions %>%
      as.data.frame() %>%
      left_join(
        data.frame(
          sample = surv_eval_dat[[i]]@originalNames,
          truth = surv_eval_dat[[i]]@actualOutcome
        ),
        by=c("sample")
      ) %>%
      mutate(subset_name = names(surv_eval_dat)[i])
  } 
) %>%
  bind_rows() %>%
  left_join(
    colData(tcga_rnaseq_data) %>% 
      as.data.frame() %>% 
      select(barcode, paper_pathologic_stage),
    by=c("sample"="barcode")
  ) %>%
  mutate(time=as.numeric(gsub("\\+","",truth)),
         event=ifelse(grepl("\\+",truth),0,1)) %>%
  group_by(subset_name,permutation,fold) %>%
  group_modify(~{
    concordance(
      truth~risk,
      data=.x,
      reverse=TRUE)$concordance %>%
      as.data.frame() %>%
      rename("cindex"=".")
  }) %>%
  group_by(subset_name)
```

#### Fig3f - C-index
```{r fig.height=5, fig.width=9}
# Plot c-index
p_cind_boxplot <- surv_risk_classifyr_cindex %>% 
  mutate(subset_name = str_replace(subset_name, "proposed_her2", "GHIST_her2")) %>%
  separate_wider_delim("subset_name", "_", names=c("data","subgroup")) %>%
  group_by(data, permutation, subgroup) %>%
  summarise(cindex = mean(cindex, na.rm=T), .groups="keep") %>%
  group_by(data) %>%
  ungroup() %>%
  left_join(
    surv_risk_classifyr_cindex %>%
      separate_wider_delim("subset_name", "_", names=c("data","subgroup"))%>% 
      group_by(data, subgroup) %>%
      summarise(cindex=mean(cindex,na.rm=T), .groups="drop") %>% 
      group_by(subgroup) %>% 
      mutate(max_ind = ifelse(cindex == max(cindex), "*","")) %>%
      select(data, subgroup, max_ind),
    by=c("data"="data","subgroup"="subgroup")
  ) %>%
  ungroup() %>% 
  mutate(data = as.factor(data)) %>% 
  mutate(subgroup=case_when(
    subgroup=="her2" ~ "HER2+",
    TRUE~ ""
  )) %>% 
  filter(!(data %in% c("RNA-Seq","RNA-Seq-STgenes", "TCGN", "EGNv2", "EGNv1"))) %>%
  mutate(data = factor(data, levels = c("GHIST", "THItoGene", "Hist2ST", "DeepPT", "DeepSpaCE", "GeneCodeR", "HisToGene", "ST-Net"))) %>% 
  ggplot() +
  aes(x=data,y=cindex, colour=data, fill=data)+
  geom_boxplot(alpha=.5) +
  facet_wrap(~subgroup, nrow=2, scales="free_x")+
  scale_fill_manual(values=dl_method_pal)+
  scale_colour_manual(values=dl_method_pal)+
  coord_flip()+
  labs(x="",y="C-index")+
  th +
  theme(legend.position = "none")

p_cind_boxplot
```

#### Figure 3g
Kaplan-Meier curves constructed using the average risk prediction from a 3-fold CV with 100 repeats within the HER2+ breast cancer subtype.

```{r fig.height=3, fig.width=17}
avg_risk_list <- lapply(
  seq_along(surv_eval_dat),
  function(i) {
    surv_eval_dat[[i]]@predictions %>%
      as.data.frame() %>%
      left_join(
        data.frame(
          sample = surv_eval_dat[[i]]@originalNames,
          truth = surv_eval_dat[[i]]@actualOutcome
        ),
        by=c("sample")
      ) %>%
      mutate(subset_name = names(surv_eval_dat)[i])
  } 
) %>%
  bind_rows() %>%
  mutate(truth = as.character(truth)) %>% 
  mutate(time=as.numeric(gsub("\\+","",truth)),
         event=ifelse(grepl("\\+",truth),0,1)) %>%
  group_by(subset_name, sample, truth, time, event) %>% 
  summarise(avg_risk = mean(risk), .groups="drop") %>%
  group_by(subset_name) %>%
  mutate(risk_cat = ifelse(avg_risk >= median(avg_risk),"high-risk","low-risk")) %>%
  split(., f=.$subset_name)

km_list <- lapply(
  seq_along(avg_risk_list),
  function(i) {
    surv_fit = survfit(Surv(time, event) ~ risk_cat,
                       data = avg_risk_list[[i]])
    
    p_km_surv <- ggsurvplot(
      fit = surv_fit,
      data = avg_risk_list[[i]],
      risk.table = FALSE,
      pval = TRUE,
      pval.size = 4,
      legend.title = "Risk Group",
      xlab = "Time (days)",
      ylab = "Survival Probability",
      title = "Kaplan-Meier Survival Curves by Risk Group"
    )
    
    # calculate logrank pvalue
    p_value <- surv_pvalue(surv_fit(
      Surv(time, event) ~ risk_cat,
      data = avg_risk_list[[i]]
    ))
  
    logrank_test <- survdiff(Surv(time, event) ~
                               risk_cat,
                             data = avg_risk_list[[i]])
    
    return(
      p_km_surv$data.survplot %>%
        mutate(pval.txt = p_value$pval.txt) %>%
        mutate(pval = p_value$pval) %>%
        mutate(subgroup = names(avg_risk_list)[i])
      )
  } 
)

# extract plot data from ggsurvplot
km_surv_plot_dat_df <- km_list %>%
  bind_rows() %>%
  separate(subgroup, c("data", "subgroup"), sep = "_") %>%
  mutate(
    subgroup = case_when(
      subgroup == "her2" ~ "HER2+",
      TRUE ~ ""
    ) %>%
      factor(., levels = c("HER2+"))
  )

th <-   theme(text=element_text(size=16),
                axis.text.x = element_text(angle = 0, hjust = 0.5),
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                panel.background = element_rect(colour = "black", size=0.7, fill=NA) )

fig_3g <- km_surv_plot_dat_df %>%
  mutate(data = str_replace(data, "proposed", "GHIST")) %>%
  filter(subgroup=="HER2+" & data!="RNA-Seq") %>%
  filter(!data %in% c("TCGN", "EGNv2", "EGNv1", "RNA-Seq-STgenes")) %>%
  mutate(
    data = factor(data,
                  levels=c("ST-Net", 
                           "HisToGene", 
                           "GeneCodeR", 
                           "DeepSpaCE", 
                           "DeepPT", 
                           "Hist2ST", 
                           "THItoGene",
                           "GHIST")
                  )
  ) %>%
  ggplot(aes(x=time/365,y=surv, col=risk_cat))+
  geom_step(linewidth = 0.5, linetype = 1)+
  geom_point(size = 4.5, shape = "+")+
  geom_text(aes(label=pval.txt, x=-Inf,y=-Inf),
            vjust=-1,hjust=-.2,
            colour="grey60", check_overlap = TRUE)+
  scale_y_continuous(limits = c(0, 1.05), expand=c(0,0))+
  scale_x_continuous(expand=c(0,0))+
  ggh4x::facet_grid2(rows=vars(subgroup), cols=vars(data), 
                     scales="free", independent = "all",switch="y")+
  theme(legend.position="none",
        strip.placement = "outside",
        strip.switch.pad.grid = unit(1, "cm"),
        axis.title.y = element_text(vjust = -15))+
  th+
  labs(x="Time (years)",y="Survival probability",col="Risk Category")

fig_3g
```
