---
title: "Supplementary figure 8"
date: "2025-4-29"
editor_options: 
  chunk_output_type: console
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Loading
```{r}
### Please download the required files for Supplementary Figure 8 from dropbox: https://www.dropbox.com/scl/fo/hs5p6te858d232wojo7dy/AH7YzDNr4lXphoqRvlp5d8w?rlkey=vvnqt96n070a136dsfn7viw2z&st=cq5fezrl&dl=0
```

## Libraries
```{r}
suppressPackageStartupMessages({
  # Load required packages
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(scran)
})

minMaxScaler <- function(x) {
  (x-min(x, na.rm=T))/(max(x, na.rm=T)-min(x, na.rm=T))
}
rename <- dplyr::rename
select <- dplyr::select
slice <- dplyr::slice

# Set ggplot theme
ggthemr_reset()
ggthemr('pale')
mycolors <- c("#999999", "#0072B2", "#E69F00", "#F0E442", "#D9B3FF", "#009E73",  
               "#D55E00", "#5D8AA8", "#CC79A7", "#56B4E9",
              "#F3B3A6", "#A5AB81", "#B2182B", "#4393C3", "#CDBE6B", 
              "#80CDC1", "#F4A582", "#BABABA", "#CCEBC5", "#DECBE4",
              "#FDDFDF", "#B3DE69", "#FDBF6F", "#CCECE6", "#FB8072")

dl_method_pal <- c(
  "THItoGene" ="#BA7FB5", 
  "DeepPT"="#80B1D2", 
  "DeepSpaCE"="#F18072",
  "ST-Net"="#8CD0C3",
  "HisToGene"="#f1c232",
  "Hist2ST"="#F9B063",
  "GeneCodeR"="#BCB9D8",
  "iStar"= "#B3D46B",
  "GHIST" = "#B2182B"
  )

mypalette <- define_palette(
  swatch = mycolors,
  gradient = c(lower = mycolors[1L], upper = mycolors[3L])
)

ggthemr(mypalette)
theme_update(panel.grid.major = element_line(linetype="dotted"))

th <-   theme(text=element_text(size=16),
                axis.text.x = element_text(angle = 0, hjust = 0.5),
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                panel.background = element_rect(colour = "black", size=0.7, fill=NA))

```


## Data

Load data for HER2ST dataset

```{r}
# Load data
# Get all expression and prediction data of the comparison spot-based SGE prediction methods
comb_pred_dat <- readRDS("./her2st_comb_pred_dat_8.rds") %>%
  mutate(model_id = ifelse(model_id =="genecoder_i500_j500",
                           "GeneCodeR", model_id))
# Get calculated metrics
pred_feat_cor <- readRDS("./her2st_pred_feat_cor_8.rds")

# Get GHIST metrics
ghist_pred_feat_cor <- readRDS("./ghist_pred_feat_cor.rds") %>%
  mutate(model_id = case_when(model_id == "Proposed" ~ "GHIST",
    TRUE ~ model_id
  ))
# Combine calculated metrics of comparison methods with GHIST
pred_feat_cor <- rbind(pred_feat_cor, ghist_pred_feat_cor)

# Get original gene expression per patch
deeppt_exprs_df <- read.csv("./processed_expr.csv")

# Highly Variable Genes
# use methods from scran (Lun, McCarthy, and Marioni 2016) to identify a set of top highly variable genes (HVGs)
dec_her2st <- modelGeneVar(deeppt_exprs_df %>%
                             select(-X) %>%
                             t())
# Get the top 10% of genes.
hv_genes <- getTopHVGs(dec_her2st, prop=.100)

# Get original gene expression from a sample method
deeppt_comb_pred_dat <- comb_pred_dat%>%
  filter(model_id=="DeepPT")
```

*comb_pred_dat* is a dataframe with the following columns:

- img_id: id of image
- model_id: the name of method predicting gene expression
- patch_id: id of image patch, containing image + patch coordinates
- pred_type: whether the evaluation metrics were calculated on the "train", "test" or 
"validation" sets
- train_fold: an id for the fold the model was trained and tested on
- gene: name of the gene that was predicted
- pred: predicted gene expression count
- exprs: ground truth gene expression count

```{r}
glimpse(comb_pred_dat)
```

*pred_feat_cor* is a dataframe with the following columns:

- img_id: id of image
- model_id: the name of method predicting gene expression
- gene: name of the gene that was predicted
- pred_type: whether the evaluation metrics were calculated on the "train", "test" or 
"validation" sets
- train_fold: an id for the fold the model was trained and tested on
- cor_pearson: Pearson correlation
- cor_spearman: Spearman correlation
- var_exprs: variance of the gene expression data the model was trained on
- var_exprs_orig: variance of the ground truth gene expression counts
- var_pred: variance of the predicted gene expression
- mean_exprs: mean of the gene expression data the model was trained on
- mean_exprs_orig: mean of the ground truth gene expression counts
- mean_pred: mean of the predicted gene expression
- rmse: root mean squared error
- mi: mutual information
- js_div: Jensen-Shannon divergence
- nrmse_range: RMSE normalised using the range of the data
- nrmse_sd: RMSE normalised using the standard deviation of data
- ssim: structural similarity index
- auc_n: original ground truth gene expression is binarised by $GE <= n$ and $GE > n$. 
auc_n represents the auc of the predicted gene expression distinguishing the binarisation of the original gene expression

```{r}
glimpse(pred_feat_cor)
```

### Load SVG
```{r}
svg_genes <- readRDS("./her2st_SVG_filter_top20.rds") 
```

```{r}
# Get GHIST predictions and reformat
ghist_comb_pred_dat <- readRDS("./ghist_comb_pred_dat.rds")
ghist_comb_pred_dat$exprs<- NULL
ghist_comb_pred_dat <- ghist_comb_pred_dat %>%
  left_join(deeppt_comb_pred_dat %>% select(img_id, gene, patch_id, exprs), 
            by = c("img_id", "gene", "patch_id"))

# Normalized gene expression used by GHIST
ghist_comb_pred_dat$exprs <- log1p(ghist_comb_pred_dat$exprs)
ghist_comb_pred_dat$model_id <- gsub("Proposed", "GHIST", ghist_comb_pred_dat$model_id)

# Combine comparison methods
comb_pred_dat_all <- rbind(comb_pred_dat, ghist_comb_pred_dat)
# Get exprs and pred gene standardization
comb_pred_dat_all$exprs <- as.numeric(unlist(comb_pred_dat_all$exprs))
comb_pred_dat_all$pred <- as.numeric(unlist(comb_pred_dat_all$pred))
comb_pred_dat_all_norm <- comb_pred_dat_all %>%
  dplyr::group_by(gene, model_id) %>%
  dplyr::mutate(
    exprs_min = min(exprs, na.rm = TRUE),
    exprs_max = max(exprs, na.rm = TRUE),
    exprs_norm = (exprs - exprs_min) / (exprs_max - exprs_min + 1e-12),
    pred_min = min(pred, na.rm = TRUE),
    pred_max = max(pred, na.rm = TRUE),
    pred_norm = (pred - pred_min) / (pred_max - pred_min + 1e-12)
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(-exprs_min, -exprs_max, -pred_min, -pred_max)

comb_pred_dat_all_norm$exprs<- as.numeric(unlist(comb_pred_dat_all_norm$exprs))
## Feature Metrics - RMSE
comb_pred_dat_all_norm_metrics <- comb_pred_dat_all_norm %>%
  # Get x & y coordinates for SSIM calculation
  separate(
    patch_id ,
    into = c("img_id", "x_y"),
    sep = "_",
    convert = TRUE,
    remove=FALSE
  ) %>%
  separate(x_y, into = c("x", "y"), sep = "x") %>%
  group_by(gene, pred_type, train_fold, model_id, img_id) %>%
  summarise(
    rmse = sqrt(sum((pred_norm - exprs_norm) ^ 2) / n()),
    .groups = "drop"
  )
```

### Supplementary Figure 8 RMSE - standardized
```{r}
rmse_df <- comb_pred_dat_all_norm_metrics%>%
  arrange(factor(model_id, levels=c("GHIST", "iStar", "THItoGene", "Hist2ST", "DeepPT", "DeepSpaCE", "GeneCodeR", "HisToGene", "ST-Net")))
rmse_df$model_id <- factor(rmse_df$model_id, levels = unique(rmse_df$model_id))

fig_rmse <- rmse_df %>%
  filter(pred_type == "test") %>%
  pivot_longer(cols=c("rmse"),
               names_to="metric",
               values_to="value") %>%
  mutate(metric = factor(metric, levels = c("rmse"))) %>%
  mutate(metric=case_when(
    metric=="rmse" ~ "RMSE",
    TRUE ~ "other"
  ) %>%
    factor(., levels=c("RMSE"))) %>%
  ggplot() +
  aes(x=model_id, y=value, fill=model_id, col=model_id) +
  geom_violin(alpha=0.2) +
  geom_boxplot(alpha=0.5, width=0.3) +
  facet_wrap(~metric, nrow=2, scales="free") +
  coord_flip()+
  scale_color_manual(values=dl_method_pal)+
  scale_fill_manual(values=dl_method_pal)+
  theme(legend.position = "none") +
  th + 
  labs(x="", y="")

fig_rmse
```

### Supplementary Figure 8 SVGs RMSE - standardized
```{r}
rmse_svg_df <- rmse_df %>%
  mutate(gene_set = "All Genes") %>%
  bind_rows(
    rmse_df %>%
      ungroup() %>%
      semi_join(svg_genes, by = c("gene", "img_id")) %>%
      mutate(gene_set = "SVGs")
  )

fig_SVG_rmse <- rmse_svg_df %>%
  filter(pred_type == "test") %>%
  filter(gene_set %in% c("SVGs")) %>%
  pivot_longer(cols=c("rmse"),
               names_to="metric",
               values_to="value") %>%
  mutate(metric = factor(metric, levels = c("rmse"))) %>%
  mutate(metric=case_when(
    metric=="rmse" ~ "RMSE",
    TRUE ~ "other"
  ) %>%
    factor(., levels=c("RMSE"))) %>%
  ggplot() +
  aes(x=model_id, y=value, fill=model_id, col=model_id) +
  geom_violin(alpha=0.2) +
  geom_boxplot(alpha=0.5, width=0.3) +
  facet_wrap(~metric, nrow=2, scales="free") +
  coord_flip()+
  scale_color_manual(values=dl_method_pal)+
  scale_fill_manual(values=dl_method_pal)+
  theme(legend.position = "none") +
  th + 
  labs(x="", y="")

fig_SVG_rmse
```

